---
title: "Manipulating the R parameter in the QS fertility model"
author: "Carl Schmertmann"
date: "01 Nov 2019"
output: pdf_document
---

## Purpose

A quick illustration of how changing the $R$ parameter 
affects the QS fertility model described in 

CP Schmertmann, 2003. *A System of Model Fertility Schedules with Graphically Intuitive Parameters*. 
**Demographic Research** Vol. 9-5 (10 Oct 2003). https://dx.doi.org/10.4054/DemRes.2003.9.5

The main idea is to find a continuous function (specifically, a restricted quadratic spline) that approximates the observed $f_x$ values.  A QS curve has four parameters:

1. $R$ : the peak level of fertility
2. $\alpha$ : the lowest age at which fertility is positive
3. $P$ : the age at which fertility reaches its peak value $R$
4. $H$ : the age $>P$ at which fertility falls to half its peak value, $R/2$.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())
graphics.off()
if (.Platform$OS.type == 'windows') windows(record=TRUE)

```


## Define the $QS(\:)$ function

$x$ is the vector of ages at which to calculate interpolated rates

```{r}
QS = function(x, R=.20,alpha=15,P=28,H=34) {
  
  w = min(.75, .25+.025*(P-alpha))
  beta = max (  min(50, 4*H-3*P) , (4*H-P)/3 )
  
  D = P-20          # delay index
  C = (P+50)/2 - H  # control index
  
  knot    = vector("numeric",5)
  knot[1] = alpha
  knot[2] = alpha + w*(P-alpha)
  knot[3] = P
  knot[4] = (P+H)/2
  knot[5] = (H+beta)/2
  
  A      = matrix(NA,5,5)
  target = vector("numeric",5)
  
  # target value at P=1
  A[1,]     = (pmax(P-knot, 0))^2
  target[1] = 1
  
  # target value at H=1/2
  A[2,]     = (pmax(H-knot, 0))^2
  target[2] = 1/2
  
  # target value at beta=0
  A[3,]     = (pmax(beta-knot, 0))^2
  target[3] = 0
  
  # target slope at P=0
  A[4,]     = 2*pmax(P-knot, 0) 
  target[4] = 0
  
  # target slope at beta=0
  A[5,]     = 2*pmax(beta-knot, 0) 
  target[5] = 0
  
  # calculate thetas
  theta = solve(A,target)
  
  tmp = (pmax( outer(x,knot,"-") , 0))^2
  
  y = (x >= alpha) * (x <= beta) * R * (tmp %*% theta)  # ASFRs
  
  return(list(R=R,alpha=alpha,P=P,H=H,
              theta=theta,knot=knot,w=w,beta=beta,delay=D,control=C,
              x=as.vector(x) ,y=as.vector(y) ))
  
} #QS
```

## The $\phi(x)$ function

The shape of a fitted QS schedule depends on the age parameters -- $\alpha$, $P$, and $H$.  In this example I'll use 
$$
(\alpha,P,H) \; = \; (15.8, 31.4, 37.1)
$$
which matches the shape for 2016 Swedish fertility rates in the Human Fertility Database. 

First produce a QS schedule over a fine grid of ages $12.25,12.75,\ldots,49.25,49.75$.

```{r}
x = seq(from=12.25, to=49.75, by=0.5)

asfr = QS(x, alpha=15.8, R=1, P=31.4, H=37.1)

plot(asfr, type='l', xlab='x', ylab='phi(x)', main='Unscaled (R=1)')
abline(v=c(15.8, 31.4, 37.1), h=c(0, 0.5, 1), lty='dashed')
```

## Direct use of R: Define the maximum fertility rate

If we change $R$ from 1 to some other value (e.g., $R=0.14$), we are changing the fertility schedule from $f(x)=1\cdot\phi(x)$ to $f(x) =R\cdot\phi(x)$, and $R$ is the new height at the mode:
```{r}

asfr = QS(x, alpha=15.8, R=0.14, P=31.4, H=37.1)

plot(asfr, type='l', xlab='x', ylab='f(x)', main='Scaled (R=0.14)')
abline(v=c(15.8, 31.4, 37.1), h=0.14*c(0, 0.5, 1), lty='dashed')
```


## Indirect use of R: Match a target TFR value

We could also choose the value of $R$ to ensure that $TFR = \int_{\alpha}^{50} f(x) dx$ matches some specific value. This value would be 
$$
R^\ast = \frac{TFR}{\int_{\alpha}^{50} \phi(x) dx}
$$

For example, if we wanted a schedule with 
$$
TFR=1.30,\,\alpha=15.8,\, P=31.4,\,H=37.1
$$
we could do the following
```{r}
phi = QS(x, alpha=15.8, R=1, P=31.4, H=37.1)

# 0.50 because that's the width of intervals on the age grid
integral_phi = sum(phi$y) * .50   

print(integral_phi)

Rstar = 1.30 / integral_phi

asfr = phi = QS(x, alpha=15.8, R=Rstar, P=31.4, H=37.1)

plot(asfr, type='l', xlab='x', ylab='f(x)', main='Scaled (TFR=1.30)')
abline(v=c(15.8, 31.4, 37.1), h=Rstar*c(0, 0.5, 1), lty='dashed')

## check
TFR = sum(asfr$y) * .50   
print(TFR)
```

## Other indirect uses of R

We could also get creative, and choose $R$ to match other features of the ASFR schedule that depend on the scaling, like

* average number of children born after age 30
* probability of having 2 children by age 35 (if we assume that births are Poisson distributed...)
* etc. 

