---
title: "Quadratic Spline Interpolation of Fertility Schedules"
author: "Carl Schmertmann"
date: "28 Nov 2018"
output:
  html_document: default
  word_document: default
---

## Purpose

Illustrates least-squares fitting of Quadratic Spline (QS) model to several examples of $_{1}f_x$ schedules from the HFD. The QS model is described in 

CP Schmertmann, 2003. *A System of Model Fertility Schedules with Graphically Intuitive Parameters*. 
**Demographic Research** Vol. 9-5 (10 Oct 2003). https://dx.doi.org/10.4054/DemRes.2003.9.5

The main idea is to find a continuous function (specifically, a restricted quadratic spline) that approximates the observed $f_x$ values.  A QS curve has four parameters:

1. $R$ : the peak level of fertility
2. $\alpha$ : the lowest age at which fertility is positive
3. $P$ : the age at which fertility reaches its peak value $R$
4. $H$ : the age $>P$ at which fertility falls to half its peak value, $R/2$.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())
graphics.off()
if (.Platform$OS.type == 'windows') windows(record=TRUE)

```

## Download and prepare several $_{1}f_x$ schedules from the HFD

```{r}
HFD  = read.table("asfrRR.txt",skip=2, header=TRUE)

HFD$Age   = 12:55  ## make numeric
HFD$sched = paste0(HFD$Code, HFD$Year) 

HFDmatrix = matrix(HFD$ASFR, nrow=44, 
                   dimnames=list(12:55, unique(HFD$sched)))


# display the first few
print( HFDmatrix[, 1:5])
```

## Identify (1) the fine grid of ages for the QS function, (2) the lowest ages in each age group for which we have nFa data, and (3) the width $n$ of each age interval

```{r}
## grid of ages for QS
delta = .25
x = seq(from= 12+delta/2, to= 56-delta/2, by=delta)

## low ends of age intervals
a = seq(from=12, to=55, by=1)

## age interval width
n = 1

## calculate translation matrix G such that G %*% QS() = grouped rates
tmp = outer( a, x, function(a,x) { 1*(x>=a)*(x <= (a+n)) })
G   = prop.table(tmp,1)

```

## Define the $QS(\:)$ function

$x$ is the vector of ages at which to calculate interpolated rates

```{r}
QS = function(R=.20,alpha=15,P=28,H=34) {
  
  w = min(.75, .25+.025*(P-alpha))
  beta = max (  min(50, 4*H-3*P) , (4*H-P)/3 )
  
  D = P-20          # delay index
  C = (P+50)/2 - H  # control index
  
  knot    = vector("numeric",5)
  knot[1] = alpha
  knot[2] = alpha + w*(P-alpha)
  knot[3] = P
  knot[4] = (P+H)/2
  knot[5] = (H+beta)/2
  
  A      = matrix(NA,5,5)
  target = vector("numeric",5)
  
  # target value at P=1
  A[1,]     = (pmax(P-knot, 0))^2
  target[1] = 1
  
  # target value at H=1/2
  A[2,]     = (pmax(H-knot, 0))^2
  target[2] = 1/2
  
  # target value at beta=0
  A[3,]     = (pmax(beta-knot, 0))^2
  target[3] = 0
  
  # target slope at P=0
  A[4,]     = 2*pmax(P-knot, 0) 
  target[4] = 0
  
  # target slope at beta=0
  A[5,]     = 2*pmax(beta-knot, 0) 
  target[5] = 0
  
  # calculate thetas
  theta = solve(A,target)
  
  tmp = (pmax( outer(x,knot,"-") , 0))^2
  
  y = (x >= alpha) * (x <= beta) * R * (tmp %*% theta)  # ASFRs
  
  return(as.vector(y) )
  
} #QS
```


```

## Create a variable to hold results

```{r}

results = data.frame(country = character(),
                     alpha   = numeric(),
                     R       = numeric(),
                     P       = numeric(),
                     H       = numeric())
```

## Calculate and plot least-squares estimates of QS parameters for 10 random HFD schedules

```{r}


#####################
selected_schedules = sample( colnames(HFDmatrix), 20)

error_formula = formula( fx ~ I( G %*% QS(R,alpha,P,H) ) ) 

for (this_schedule in selected_schedules) {
  
    fx <- HFDmatrix[, this_schedule]  

    ## INITIAL ESTIMATES OF PARAMETERS using MAC for P
    midage <- 12:55 + 0.5
    mac   <- sum(fx*midage)/sum(fx)
    init_param <- c(alpha=15,R=max(fx),P=mac, H=(mac+50)/2)
    
    
    ## find optimal R, P & H params for QS
    this.result <- nls(error_formula, 
                       start=init_param,
                       trace=TRUE)
    
    results = rbind(results,
                    data.frame(sched  = this_schedule, 
                               alpha  = round(coef(this.result)['alpha'],2),
                               R      = round(coef(this.result)['R'],2),
                               P      = round(coef(this.result)['P'],2),
                               H      = round(coef(this.result)['H'],2),2)
                    )
    
    
    #-------- plot --------------
    yhat   = QS(alpha=15, R=this.result$par['R'], P=this.result$par['P'], H=this.result$par['H'])


    plot( x, yhat, type='o', main=this_schedule, pch=16, cex=.60,
          ylim=range(0,yhat,fx))
    
    L = seq(12,55,1)
    segments(L, fx, L+1,  fx, col=2, lwd=2)
    segments(L,     0, L  , fx, col=2, lwd=2)
    segments(L+1,   0, L+1, fx, col=2, lwd=2)

} # for this_schedule
```

## Display the parameter estimates that correspond to the graphs

```{r}
rownames(results) = NULL
results 